---
title: "GLEAM Seurat v4/v5 input guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GLEAM Seurat v4/v5 input guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7.2,
  fig.height = 5.2,
  out.width = "88%",
  dpi = 120
)
library(GLEAM)
```

## Why this guide

`score_signature()` supports Seurat v4 and v5 with the same top-level API.
This guide demonstrates:

1. Standard Seurat preprocessing before scoring
2. v5-style `layer` usage
3. v4-compatible `slot` fallback
4. Reduction extraction and embedding-based signature display

## 1) Build a Seurat object and run standard workflow

```{r build-seurat, eval=requireNamespace("Seurat", quietly = TRUE), fig.width=7.5, fig.height=5.4}
library(Seurat)

ifnb_path <- system.file("extdata", "full_examples", "ifnb_seurat.rds", package = "GLEAM")
if (ifnb_path == "") ifnb_path <- file.path("inst", "extdata", "full_examples", "ifnb_seurat.rds")
seu <- readRDS(ifnb_path)
if (ncol(seu) > 5000) seu <- subset(seu, cells = colnames(seu)[1:5000])

if (!"pca" %in% names(seu@reductions)) {
  seu <- NormalizeData(seu)
  seu <- FindVariableFeatures(seu)
  seu <- ScaleData(seu)
  seu <- RunPCA(seu)
}
if (!"umap" %in% names(seu@reductions)) {
  seu <- FindNeighbors(seu, dims = 1:20)
  seu <- FindClusters(seu, resolution = 0.4)
  seu <- RunUMAP(seu, dims = 1:20)
}
if (!"tsne" %in% names(seu@reductions)) {
  seu <- RunTSNE(seu, dims = 1:20)
}

dim(seu)
seu
table(Idents(seu))
DimPlot(seu, reduction = "umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE)
```

## 2) Seurat v5-preferred path (`layer`)

```{r resolve-hallmark, eval=requireNamespace("Seurat", quietly = TRUE)}
map_geneset_to_expr <- function(gs, expr_genes) {
  expr_genes <- as.character(expr_genes)
  expr_upper <- toupper(expr_genes)
  mapped <- lapply(gs, function(g) {
    idx <- match(toupper(unique(as.character(g))), expr_upper, nomatch = 0L)
    unique(expr_genes[idx[idx > 0L]])
  })
  mapped[vapply(mapped, length, integer(1)) > 0L]
}

fallback_signatures <- function(expr_genes) {
  g <- unique(as.character(expr_genes))
  g <- g[!is.na(g) & nzchar(g)]
  k <- min(30L, max(3L, floor(length(g) / 4)))
  idx1 <- seq_len(min(k, length(g)))
  idx2 <- seq.int(max(1L, length(g) - k + 1L), length(g))
  list(
    Signature_A = g[idx1],
    Signature_B = g[idx2]
  )
}

gs_for_score <- NULL
for (sp in c("human", "mouse")) {
  gs_try <- try(get_geneset("hallmark", source = "builtin", species = sp), silent = TRUE)
  if (inherits(gs_try, "try-error")) next
  gs_try <- map_geneset_to_expr(gs_try, rownames(seu))
  gs_try <- gs_try[vapply(gs_try, length, integer(1)) >= 3L]
  if (length(gs_try) > 0L) {
    gs_for_score <- gs_try
    break
  }
}
if (is.null(gs_for_score) || length(gs_for_score) == 0L) {
  message("[GLEAM] No matched Hallmark signatures; using object-derived fallback signatures.")
  gs_for_score <- fallback_signatures(rownames(seu))
}
```

```{r seurat-v5-layer, eval=requireNamespace("Seurat", quietly = TRUE)}
sc_v5 <- score_signature(
  object = seu,
  geneset = gs_for_score,
  geneset_source = "list",
  seurat = TRUE,
  assay = "RNA",
  layer = "data",
  slot = "data",
  method = "rank",
  min_genes = 3
)
```

## 3) Seurat v4-compatible path (`slot` fallback)

```{r seurat-v4-slot, eval=requireNamespace("Seurat", quietly = TRUE)}
sc_v4 <- score_signature(
  object = seu,
  geneset = gs_for_score,
  geneset_source = "list",
  seurat = TRUE,
  assay = "RNA",
  layer = NULL,
  slot = "data",
  method = "rank",
  min_genes = 3
)
```

## 4) Reduction extraction and signature visualization

```{r reduction-and-plot, eval=requireNamespace("Seurat", quietly = TRUE)}
emb_umap <- extract_embedding(object = seu, reduction = "umap")
head(emb_umap)

top_sig <- rownames(sc_v5$score)[1]
plot_embedding_score(sc_v5, pathway = top_sig, object = seu, reduction = "umap")
plot_embedding_score(sc_v5, pathway = top_sig, object = seu, reduction = "pca")
plot_embedding_score(sc_v5, pathway = top_sig, object = seu, reduction = "tsne")
```
