---
title: "GLEAM full spatial transcriptomics workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GLEAM full spatial transcriptomics workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7.4,
  fig.height = 5.3,
  out.width = "88%",
  dpi = 120
)
library(GLEAM)
```

## 1) Spatial object construction

Although this is a spatial workflow, running a standard Seurat pipeline first
helps align signature interpretation across region labels, learned clusters, and
low-dimensional embeddings.

```{r full-spatial-object, eval=requireNamespace("Seurat", quietly = TRUE), fig.width=7.6, fig.height=5.6, out.width="92%"}
library(Seurat)

a1_path <- system.file("extdata", "full_examples", "stxBrain_anterior1_seurat.rds", package = "GLEAM")
p1_path <- system.file("extdata", "full_examples", "stxBrain_posterior1_seurat.rds", package = "GLEAM")
if (a1_path == "") a1_path <- file.path("inst", "extdata", "full_examples", "stxBrain_anterior1_seurat.rds")
if (p1_path == "") p1_path <- file.path("inst", "extdata", "full_examples", "stxBrain_posterior1_seurat.rds")

sp_a1 <- readRDS(a1_path)
sp_p1 <- readRDS(p1_path)
sp_seu <- merge(sp_a1, y = sp_p1, add.cell.ids = c("anterior", "posterior"))
if (ncol(sp_seu) > 8000) sp_seu <- subset(sp_seu, cells = colnames(sp_seu)[1:8000])

sp_md <- sp_seu@meta.data
if (!"sample" %in% colnames(sp_md)) sp_md$sample <- if ("orig.ident" %in% colnames(sp_md)) as.character(sp_md$orig.ident) else "sample_1"
if (!"group" %in% colnames(sp_md)) sp_md$group <- ifelse(grepl("anterior", sp_md$sample, ignore.case = TRUE), "anterior", "posterior")
if (!"region" %in% colnames(sp_md)) sp_md$region <- if ("seurat_clusters" %in% colnames(sp_md)) paste0("cluster_", sp_md$seurat_clusters) else sp_md$group
if (!"section" %in% colnames(sp_md)) sp_md$section <- sp_md$sample
sp_seu@meta.data <- sp_md

if (!"pca" %in% names(sp_seu@reductions)) {
  sp_seu <- NormalizeData(sp_seu)
  sp_seu <- FindVariableFeatures(sp_seu)
  sp_seu <- ScaleData(sp_seu)
  sp_seu <- RunPCA(sp_seu)
}
ElbowPlot(sp_seu, ndims = 30)
if (!"umap" %in% names(sp_seu@reductions)) {
  sp_seu <- FindNeighbors(sp_seu, dims = 1:20)
  sp_seu <- FindClusters(sp_seu, resolution = 0.5)
  sp_seu <- RunUMAP(sp_seu, dims = 1:20)
}
if (!"tsne" %in% names(sp_seu@reductions)) {
  sp_seu <- RunTSNE(sp_seu, dims = 1:20)
}
if (!"pseudotime" %in% colnames(sp_seu@meta.data)) {
  sp_seu$pseudotime <- rank(Embeddings(sp_seu, "pca")[, 1], ties.method = "average") / ncol(sp_seu)
}
if (!"lineage" %in% colnames(sp_seu@meta.data)) {
  sp_seu$lineage <- as.character(sp_seu$region)
}

# Show object context before scoring
dim(sp_seu)
sp_seu
head(sp_seu@meta.data[, intersect(c("sample", "group", "region", "section", "x", "y", "imagecol", "imagerow"), colnames(sp_seu@meta.data))])
table(Idents(sp_seu))
table(cluster = Idents(sp_seu), region = sp_seu$region)
DimPlot(sp_seu, reduction = "umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE, pt.size = 0.35)
DimPlot(sp_seu, reduction = "umap", group.by = "region", pt.size = 0.35)
DimPlot(sp_seu, reduction = "tsne", group.by = "section", pt.size = 0.35)
```

## 2) Geneset scoring in spatial context

```{r full-spatial-score, eval=requireNamespace("Seurat", quietly = TRUE)}
sp_genes <- rownames(sp_seu)
half_n <- max(30, floor(length(sp_genes) / 2))
idx_a <- seq_len(min(half_n, length(sp_genes)))
idx_b <- seq(from = min(half_n + 1, length(sp_genes)), to = length(sp_genes))
gs <- list(
  Spatial_signature_A = unique(sp_genes[idx_a])[1:min(30, length(unique(sp_genes[idx_a])))],
  Spatial_signature_B = unique(sp_genes[idx_b])[1:min(30, length(unique(sp_genes[idx_b])))]
)

sp <- score_signature(
  object = sp_seu,
  geneset = gs,
  geneset_source = "list",
  seurat = TRUE,
  assay = "RNA",
  layer = "data",
  slot = "data",
  method = "rank",
  min_genes = 3
)
```

## 3) Spatial visualization (true spatial context)

```{r full-spatial-map, eval=requireNamespace("Seurat", quietly = TRUE), fig.width=7.6, fig.height=5.6, out.width="92%"}
coords <- if (all(c("x", "y") %in% colnames(sp_seu@meta.data))) {
  data.frame(x = sp_seu@meta.data$x, y = sp_seu@meta.data$y, row.names = rownames(sp_seu@meta.data))
} else if (all(c("imagecol", "imagerow") %in% colnames(sp_seu@meta.data))) {
  data.frame(x = sp_seu@meta.data$imagecol, y = sp_seu@meta.data$imagerow, row.names = rownames(sp_seu@meta.data))
} else {
  emb <- Embeddings(sp_seu, reduction = "pca")[, 1:2, drop = FALSE]
  data.frame(x = emb[, 1], y = emb[, 2], row.names = rownames(emb))
}

# Slice/tissue-like background (replace with true tissue image when available)
tissue_bg <- as.raster(matrix(colorRampPalette(c("#f8f5ea", "#e8dcc4", "#d2b48c"))(256), nrow = 16))

plot_spatial_score(sp, pathway = rownames(sp$score)[1], coords = coords, image = tissue_bg)
plot_spatial_score(sp, pathway = rownames(sp$score)[2], coords = coords, split.by = "section", image = tissue_bg)
plot_spatial_multi(sp, pathways = rownames(sp$score)[1:4], coords = coords)
```

## 4) Spatial differential analysis

```{r full-spatial-diff, eval=requireNamespace("Seurat", quietly = TRUE), fig.width=7.6, fig.height=5.6, out.width="92%"}
# region vs region
res_region <- test_signature(sp, region = "region", level = "region", method = "wilcox")

# group vs group within region
res_group_region <- test_signature(
  sp,
  region = "region",
  group = "group",
  sample = "sample",
  level = "sample_region",
  method = "wilcox"
)

top_sp <- res_group_region$table$pathway[order(res_group_region$table$p_adj)][1]
plot_spatial_score(sp, pathway = top_sp, coords = coords, image = tissue_bg, split.by = "region")
plot_spatial_compare(res_group_region)
```

## 5) Optional embedding linkage and trajectory-like display

```{r full-spatial-embedding-link, eval=requireNamespace("Seurat", quietly = TRUE), fig.width=7.6, fig.height=5.6, out.width="92%"}
plot_embedding_score(sp, pathway = rownames(sp$score)[1], object = sp_seu, reduction = "umap")
plot_pseudotime_score(sp, pathway = rownames(sp$score)[1], pseudotime = "pseudotime", lineage = "lineage")
```
