---
title: "GLEAM full spatial transcriptomics workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GLEAM full spatial transcriptomics workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 9.8,
  fig.height = 7.2,
  out.width = "96%",
  dpi = 180
)
pkg_root <- if (file.exists("DESCRIPTION")) "." else if (file.exists(file.path("..", "DESCRIPTION"))) ".." else NULL
if (!is.null(pkg_root) && requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(pkg_root, export_all = FALSE, quiet = TRUE)
} else if (requireNamespace("GLEAM", quietly = TRUE)) {
  library(GLEAM)
} else {
  stop("Need installed 'GLEAM' or package 'pkgload' to render this vignette.")
}

.stratified_keep <- GLEAM:::.stratified_keep
.prep_spatial_object_matrix <- GLEAM:::.prep_spatial_object_matrix
.map_geneset_to_expr <- GLEAM:::.map_geneset_to_expr
.fallback_signatures <- GLEAM:::.fallback_signatures
```

## 1) Spatial object construction

This workflow loads Seurat spatial objects, then combines expression and metadata
at matrix level to avoid image-slot compatibility breakpoints across Seurat
versions while retaining true spatial coordinate analysis.

```{r full-spatial-fallback-figs, eval=!requireNamespace("Seurat", quietly = TRUE), echo=FALSE, fig.width=9.8, fig.height=7.2, out.width="96%"}
cat("Seurat is not available in this rendering environment. Showing representative pre-rendered spatial workflow figures.\n\n")
fig_root_candidates <- c("man/figures", "../man/figures", "../../man/figures")
fig_root <- fig_root_candidates[file.exists(fig_root_candidates)][1]
if (is.na(fig_root)) fig_root <- "man/figures"
knitr::include_graphics(c(
  file.path(fig_root, "spatial_slice_signature.png"),
  file.path(fig_root, "signature_dotbar_compare.png")
))
```

```{r full-spatial-object, eval=requireNamespace("Seurat", quietly = TRUE), fig.width=9.8, fig.height=7.2, out.width="96%"}
library(Seurat)

a1_path <- system.file("extdata", "full_examples", "stxBrain_anterior1_seurat.rds", package = "GLEAM")
p1_path <- system.file("extdata", "full_examples", "stxBrain_posterior1_seurat.rds", package = "GLEAM")
if (a1_path == "") a1_path <- file.path("inst", "extdata", "full_examples", "stxBrain_anterior1_seurat.rds")
if (p1_path == "") p1_path <- file.path("inst", "extdata", "full_examples", "stxBrain_posterior1_seurat.rds")

sp_a1 <- readRDS(a1_path)
sp_p1 <- readRDS(p1_path)
if (requireNamespace("SeuratObject", quietly = TRUE)) {
  sp_a1 <- tryCatch(SeuratObject::UpdateSeuratObject(sp_a1), error = function(e) sp_a1)
  sp_p1 <- tryCatch(SeuratObject::UpdateSeuratObject(sp_p1), error = function(e) sp_p1)
}
sp1 <- .prep_spatial_object_matrix(sp_a1, "anterior")
sp2 <- .prep_spatial_object_matrix(sp_p1, "posterior")
common_genes <- intersect(rownames(sp1$expr), rownames(sp2$expr))
sp_expr <- cbind(sp1$expr[common_genes, , drop = FALSE], sp2$expr[common_genes, , drop = FALSE])
sp_md <- rbind(sp1$meta[colnames(sp1$expr), , drop = FALSE], sp2$meta[colnames(sp2$expr), , drop = FALSE])
sp_md <- sp_md[colnames(sp_expr), , drop = FALSE]
if (ncol(sp_expr) > 8000) {
  keep <- .stratified_keep(
    meta = sp_md,
    n_target = 8000,
    strata_candidates = c("sample", "section", "group", "region", "seurat_clusters")
  )
  sp_expr <- sp_expr[, keep, drop = FALSE]
  sp_md <- sp_md[keep, , drop = FALSE]
}
if (!"group" %in% colnames(sp_md)) sp_md$group <- ifelse(grepl("anterior", sp_md$sample, ignore.case = TRUE), "anterior", "posterior")
if (!"region" %in% colnames(sp_md)) sp_md$region <- if ("seurat_clusters" %in% colnames(sp_md)) paste0("cluster_", sp_md$seurat_clusters) else sp_md$group
if (!"section" %in% colnames(sp_md)) sp_md$section <- sp_md$sample
if (!all(c("x", "y") %in% colnames(sp_md))) {
  if (all(c("array_col", "array_row") %in% colnames(sp_md))) {
    sp_md$x <- sp_md$array_col
    sp_md$y <- sp_md$array_row
  } else {
    n <- nrow(sp_md)
    sp_md$x <- seq_len(n)
    sample_vec <- if ("sample" %in% colnames(sp_md)) sp_md$sample else rep("sample_1", n)
    sp_md$y <- as.numeric(as.factor(sample_vec))
  }
}
if (!"pseudotime" %in% colnames(sp_md)) {
  sp_md$pseudotime <- rank(sp_md$x, ties.method = "average") / nrow(sp_md)
}
if (!"lineage" %in% colnames(sp_md)) {
  sp_md$lineage <- as.character(sp_md$region)
}

# Show object context before scoring
dim(sp_a1); sp_a1
dim(sp_p1); sp_p1
head(sp_md[, intersect(c("sample", "group", "region", "section", "x", "y", "imagecol", "imagerow"), colnames(sp_md))])
if ("umap" %in% names(sp_a1@reductions)) {
  DimPlot(sp_a1, reduction = "umap", group.by = "orig.ident", pt.size = 0.35)
}
```

## 2) Geneset scoring in spatial context

```{r full-spatial-score, eval=requireNamespace("Seurat", quietly = TRUE)}
sp_genes <- rownames(sp_expr)
half_n <- max(30, floor(length(sp_genes) / 2))
idx_a <- seq_len(min(half_n, length(sp_genes)))
idx_b <- seq(from = min(half_n + 1, length(sp_genes)), to = length(sp_genes))
gs <- list(
  Spatial_signature_A = unique(sp_genes[idx_a])[1:min(30, length(unique(sp_genes[idx_a])))],
  Spatial_signature_B = unique(sp_genes[idx_b])[1:min(30, length(unique(sp_genes[idx_b])))]
)

sp <- score_signature(
  expr = sp_expr,
  meta = sp_md,
  geneset = gs,
  geneset_source = "list",
  seurat = FALSE,
  method = "rank",
  min_genes = 3
)
gs_slice <- .map_geneset_to_expr(gs, rownames(sp_a1), min_genes = 3)
if (length(gs_slice) == 0L) {
  gs_slice <- .fallback_signatures(rownames(sp_a1), max_genes = 30, min_genes = 3)
}

sp_slice <- score_signature(
  object = sp_a1,
  geneset = gs_slice,
  geneset_source = "list",
  seurat = TRUE,
  layer = "counts",
  slot = "counts",
  method = "rank",
  min_genes = 3
)
```

## 3) Spatial visualization (native slice + combined coordinates)

```{r full-spatial-map, eval=requireNamespace("Seurat", quietly = TRUE), fig.width=11.2, fig.height=8.2, out.width="100%"}
coords <- data.frame(x = sp_md$x, y = sp_md$y, row.names = rownames(sp_md))

# Native Seurat slice/tissue rendering (recommended)
if (requireNamespace("Seurat", quietly = TRUE)) {
  suppressWarnings(
    try(
      Seurat::SpatialDimPlot(sp_a1, group.by = "orig.ident", pt.size.factor = 1.6),
      silent = TRUE
    )
  )
}

slice_pathway <- rownames(sp_slice$score)[1]
sig_col <- paste0("GLEAM_signature_", gsub("[^A-Za-z0-9_]+", "_", slice_pathway))
slice_vals <- rep(NA_real_, ncol(sp_a1))
names(slice_vals) <- colnames(sp_a1)
slice_cells <- intersect(colnames(sp_a1), colnames(sp_slice$score))
slice_vals[slice_cells] <- as.numeric(sp_slice$score[slice_pathway, slice_cells])
sp_a1[[sig_col]] <- slice_vals

native_sp <- plot_spatial_score(sp_slice, pathway = slice_pathway, object = sp_a1)
native_sp

# Combined-sample coordinate view (for across-slice comparative context)
tissue_bg <- as.raster(matrix(colorRampPalette(c("#f8f5ea", "#e8dcc4", "#d2b48c"))(256), nrow = 16))

plot_spatial_score(sp, pathway = rownames(sp$score)[1], coords = coords, image = tissue_bg)
plot_spatial_score(sp, pathway = rownames(sp$score)[2], coords = coords, split.by = "section", image = tissue_bg)
plot_spatial_multi(sp, pathways = rownames(sp$score)[1:4], coords = coords)
```

## 4) Spatial differential analysis

```{r full-spatial-diff, eval=requireNamespace("Seurat", quietly = TRUE), fig.width=11.4, fig.height=8.4, out.width="100%"}
# region vs region
res_region <- test_signature(sp, region = "region", level = "region", method = "wilcox")

# group vs group within region
res_group_region <- test_signature(
  sp,
  region = "region",
  group = "group",
  sample = "sample",
  level = "sample_region",
  method = "wilcox"
)

top_sp <- res_group_region$table$pathway[order(res_group_region$table$p_adj)][1]
plot_spatial_score(sp, pathway = top_sp, coords = coords, image = tissue_bg, split.by = "region")
plot_spatial_compare(res_group_region)
```

## 5) Optional embedding linkage and trajectory-like display

```{r full-spatial-embedding-link, eval=requireNamespace("Seurat", quietly = TRUE), fig.width=10.0, fig.height=7.2, out.width="96%"}
emb2 <- as.matrix(sp_md[, c("x", "y"), drop = FALSE])
rownames(emb2) <- rownames(sp_md)
colnames(emb2) <- c("Spatial_1", "Spatial_2")
plot_embedding_score(sp, pathway = rownames(sp$score)[1], embedding = emb2, reduction = "umap")
plot_pseudotime_score(sp, pathway = rownames(sp$score)[1], pseudotime = "pseudotime", lineage = "lineage")
```
