---
title: "GLEAM full scRNA-seq workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GLEAM full scRNA-seq workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7.4,
  fig.height = 5.3,
  out.width = "88%",
  dpi = 120
)
library(GLEAM)
```

## 1) Build a Seurat object with standard preprocessing

For realistic scRNA analysis, signature scoring is most useful after a standard
cell-state workflow (normalization, variable feature selection, dimensionality
reduction, neighbor graph construction, and clustering).

```{r full-scrna-object, eval=requireNamespace("Seurat", quietly = TRUE), fig.width=7.6, fig.height=5.6, out.width="92%"}
library(Seurat)
data("pbmc_medium_matrix", package = "GLEAM")
data("pbmc_medium_meta", package = "GLEAM")

seu <- CreateSeuratObject(counts = pbmc_medium_matrix, meta.data = pbmc_medium_meta)

# Basic QC context for tutorial reproducibility.
if (any(grepl("^MT-", rownames(seu)))) {
  seu[["percent.mt"]] <- PercentageFeatureSet(seu, pattern = "^MT-")
}
qc_cols <- intersect(c("nFeature_RNA", "nCount_RNA", "percent.mt"), colnames(seu@meta.data))
summary(seu@meta.data[, qc_cols, drop = FALSE])

# Minimal, conventional Seurat preprocessing path.
seu <- NormalizeData(seu)
seu <- FindVariableFeatures(seu)
seu <- ScaleData(seu)
seu <- RunPCA(seu)
ElbowPlot(seu, ndims = 30)
seu <- FindNeighbors(seu, dims = 1:20)
seu <- FindClusters(seu, resolution = 0.5)
seu <- RunUMAP(seu, dims = 1:20)
seu <- RunTSNE(seu, dims = 1:20)

# Show object context before downstream analysis
dim(seu)
seu
head(seu@meta.data[, c("sample", "group", "celltype", "pseudotime", "lineage")])
table(Idents(seu))
DimPlot(seu, reduction = "umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE, pt.size = 0.35)
DimPlot(seu, reduction = "umap", group.by = "celltype", pt.size = 0.35)
DimPlot(seu, reduction = "tsne", group.by = "group", pt.size = 0.35)
```

## 2) Built-in geneset scoring + optional custom scoring

```{r full-scrna-score, eval=requireNamespace("Seurat", quietly = TRUE)}
gs_h <- get_geneset("hallmark", source = "builtin", species = "human")

sc <- score_signature(
  object = seu,
  geneset = gs_h,
  geneset_source = "list",
  seurat = TRUE,
  assay = "RNA",
  layer = "data",
  slot = "data",
  method = "ensemble"
)

custom <- list(IFN_custom = c("STAT1", "IRF1", "ISG15", "IFIT3"))
sc_custom <- score_signature(object = seu, geneset = custom, geneset_source = "list", seurat = TRUE, method = "mean")
```

## 3) Differential analysis after scoring

```{r full-scrna-diff, eval=requireNamespace("Seurat", quietly = TRUE)}
# cell-level exploratory
res_cell <- test_signature(sc, group = "group", level = "cell", method = "wilcox")

# sample-level
res_sample <- test_signature(sc, group = "group", sample = "sample", level = "sample", method = "wilcox")

# pseudobulk sample + celltype
res_pb <- test_signature(sc, group = "group", sample = "sample", celltype = "celltype", level = "pseudobulk", method = "wilcox")

# within-celltype
res_wct <- compare_groups_within_celltype(sc, group = "group", celltype = "celltype", target_celltype = "CD8_T", sample = "sample", level = "sample")
```

## 4) Embedding and Seurat-style visualizations

```{r full-scrna-plots, eval=requireNamespace("Seurat", quietly = TRUE), fig.width=7.6, fig.height=5.6, out.width="92%"}
top_pw <- res_pb$table$pathway[order(res_pb$table$p_adj)][1]
plot_embedding_score(sc, pathway = top_pw, object = seu, reduction = "umap")
plot_embedding_score(sc, pathway = top_pw, object = seu, reduction = "pca")
plot_embedding_score(sc, pathway = top_pw, object = seu, reduction = "tsne")

plot_dot(sc, by = c("group", "celltype"))
plot_dot_bar(sc, by = c("group", "celltype"))
plot_violin(sc, pathway = rownames(sc$score)[1], group = "group")
plot_split_violin(sc, pathway = rownames(sc$score)[1], x = "celltype", split.by = "group")
plot_ridge(sc, pathway = rownames(sc$score)[1], group = "celltype")
plot_pseudobulk_box(sc, pathway = rownames(sc$score)[1], group = "group", sample = "sample", celltype = "celltype")
plot_volcano(res_pb)
```

## 5) Trajectory / pseudotime integration

```{r full-scrna-trajectory, eval=requireNamespace("Seurat", quietly = TRUE), fig.width=7.6, fig.height=5.6, out.width="92%"}
# Generic metadata-driven trajectory fields
traj_res <- test_signature_trajectory(sc, pseudotime = "pseudotime", lineage = "lineage", method = "spearman")
plot_pseudotime_score(sc, pathway = rownames(sc$score)[1], pseudotime = "pseudotime", lineage = "lineage")
plot_trajectory_score(sc, pathway = rownames(sc$score)[1], object = seu, reduction = "umap")
```

### Optional Monocle2 / Monocle3 / slingshot object pathways

```{r, eval=FALSE}
# Monocle2 object (package monocle)
# traj_res_m2 <- test_signature_trajectory(sc, pseudotime = cds_monocle2, lineage = cds_monocle2, method = "spearman")

# Monocle3 object (package monocle3)
# traj_res_m3 <- test_signature_trajectory(sc, pseudotime = cds_monocle3, lineage = "lineage", method = "spearman")

# slingshot object
# traj_res_sl <- test_signature_trajectory(sc, pseudotime = sds, lineage = sds, method = "spearman")
```

## 6) Export score tables and compare methods

```{r full-scrna-export, eval=requireNamespace("Seurat", quietly = TRUE)}
long_tbl <- pivot_scores_long(sc)
sum_tbl <- summarize_scores(sc, by = c("sample", "group"), fun = "mean")
export_scores(sc, "gleam_scores_scrna.csv", format = "csv", include_meta = TRUE)

head(long_tbl)
head(sum_tbl)
```
