---
title: "GLEAM visualization parameters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GLEAM visualization parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(GLEAM)
```

## Shared readability controls

GLEAM plots return `ggplot` objects and support centralized styling through `gleam_theme()` / `apply_gleam_theme()`.

```{r, eval=FALSE}
p <- plot_violin(sc, pathway = "HALLMARK_INTERFERON_GAMMA_RESPONSE", group = "group")
apply_gleam_theme(
  p,
  base_size = 14,
  title_size = 18,
  axis_text_size = 12,
  legend_text_size = 11,
  font_family = "sans",
  title_color = "#1f2937",
  axis_text_color = "#111827"
)
```

## Palette controls

```{r}
list_palettes()
get_palette("gleam_discrete", n = 6)
get_palette(c("#264653", "#2a9d8f", "#e9c46a", "#f4a261", "#e76f51"), n = 5)
```

## Common plotting arguments

- `pathway`: single pathway to visualize (violin, embedding, spatial).
- `by`: aggregation group keys (`plot_dot`, `plot_dot_bar`, `plot_heatmap`).
- `group` / `split.by`: grouped/split displays.
- `reduction`: embedding selection (`umap`, `pca`, `tsne`) when Seurat object is supplied.
- `point_size`, `alpha`: readability and density controls.
- `palette`: named or custom vectors.
- Spatial-specific:
  - `coords`: x/y coordinates
  - `image`: tissue/slice background (optional)
  - `split.by`: sample/section panels

