---
title: "GLEAM Seurat v4/v5 input guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GLEAM Seurat v4/v5 input guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7.2,
  fig.height = 5.2,
  out.width = "88%",
  dpi = 120
)
library(GLEAM)
```

## Why this guide

`score_signature()` supports Seurat v4 and v5 with the same top-level API.
This guide demonstrates:

1. Standard Seurat preprocessing before scoring
2. v5-style `layer` usage
3. v4-compatible `slot` fallback
4. Reduction extraction and embedding-based signature display

## 1) Build a Seurat object and run standard workflow

```{r build-seurat, eval=requireNamespace("Seurat", quietly = TRUE), fig.width=7.5, fig.height=5.4}
library(Seurat)
data("pbmc_medium_matrix", package = "GLEAM")
data("pbmc_medium_meta", package = "GLEAM")

seu <- CreateSeuratObject(counts = pbmc_medium_matrix, meta.data = pbmc_medium_meta)
seu <- NormalizeData(seu)
seu <- FindVariableFeatures(seu)
seu <- ScaleData(seu)
seu <- RunPCA(seu)
seu <- FindNeighbors(seu, dims = 1:20)
seu <- FindClusters(seu, resolution = 0.4)
seu <- RunUMAP(seu, dims = 1:20)
seu <- RunTSNE(seu, dims = 1:20)

dim(seu)
seu
table(Idents(seu))
DimPlot(seu, reduction = "umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE)
```

## 2) Seurat v5-preferred path (`layer`)

```{r seurat-v5-layer, eval=requireNamespace("Seurat", quietly = TRUE)}
sc_v5 <- score_signature(
  object = seu,
  geneset = "hallmark",
  geneset_source = "builtin",
  species = "human",
  seurat = TRUE,
  assay = "RNA",
  layer = "data",
  slot = "data",
  method = "rank",
  min_genes = 3
)
```

## 3) Seurat v4-compatible path (`slot` fallback)

```{r seurat-v4-slot, eval=requireNamespace("Seurat", quietly = TRUE)}
sc_v4 <- score_signature(
  object = seu,
  geneset = "hallmark",
  geneset_source = "builtin",
  species = "human",
  seurat = TRUE,
  assay = "RNA",
  layer = NULL,
  slot = "data",
  method = "rank",
  min_genes = 3
)
```

## 4) Reduction extraction and signature visualization

```{r reduction-and-plot, eval=requireNamespace("Seurat", quietly = TRUE)}
emb_umap <- extract_embedding(object = seu, reduction = "umap")
head(emb_umap)

top_sig <- rownames(sc_v5$score)[1]
plot_embedding_score(sc_v5, pathway = top_sig, object = seu, reduction = "umap")
plot_embedding_score(sc_v5, pathway = top_sig, object = seu, reduction = "pca")
plot_embedding_score(sc_v5, pathway = top_sig, object = seu, reduction = "tsne")
```
