---
title: "GLEAM quick start: spatial Seurat workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GLEAM quick start: spatial Seurat workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7.2,
  fig.height = 5.2,
  out.width = "85%",
  dpi = 120
)
library(GLEAM)
```

## Goal

This quick start demonstrates:

1. Spatial Seurat input
2. Standard preprocessing (normalization, HVG, PCA, neighbors, clustering, UMAP)
3. Signature scoring on merged `stxBrain_anterior1` + `stxBrain_posterior1`
4. Region/sample-aware differential testing
5. Spatial score mapping with explicit slice/tissue context

## 1) Load/create spatial-capable Seurat object

We keep the same conventional preprocessing logic used in scRNA analysis
(normalization, PCA, neighbors, clustering, UMAP) so spatial signature patterns
can be interpreted together with inferred spot states.

```{r spatial-seurat-object, eval=requireNamespace("Seurat", quietly = TRUE), fig.width=7.5, fig.height=5.4, out.width="90%"}
library(Seurat)

a1_path <- system.file("extdata", "full_examples", "stxBrain_anterior1_seurat.rds", package = "GLEAM")
p1_path <- system.file("extdata", "full_examples", "stxBrain_posterior1_seurat.rds", package = "GLEAM")
if (a1_path == "") a1_path <- file.path("inst", "extdata", "full_examples", "stxBrain_anterior1_seurat.rds")
if (p1_path == "") p1_path <- file.path("inst", "extdata", "full_examples", "stxBrain_posterior1_seurat.rds")

sp_a1 <- readRDS(a1_path)
sp_p1 <- readRDS(p1_path)
sp_seu <- merge(sp_a1, y = sp_p1, add.cell.ids = c("anterior", "posterior"))
if (ncol(sp_seu) > 6000) sp_seu <- subset(sp_seu, cells = colnames(sp_seu)[1:6000])

sp_md <- sp_seu@meta.data
if (!"sample" %in% colnames(sp_md)) sp_md$sample <- if ("orig.ident" %in% colnames(sp_md)) as.character(sp_md$orig.ident) else "sample_1"
if (!"group" %in% colnames(sp_md)) sp_md$group <- ifelse(grepl("anterior", sp_md$sample, ignore.case = TRUE), "anterior", "posterior")
if (!"region" %in% colnames(sp_md)) sp_md$region <- if ("seurat_clusters" %in% colnames(sp_md)) paste0("cluster_", sp_md$seurat_clusters) else sp_md$group
sp_seu@meta.data <- sp_md

if (!"pca" %in% names(sp_seu@reductions)) {
  sp_seu <- NormalizeData(sp_seu)
  sp_seu <- FindVariableFeatures(sp_seu)
  sp_seu <- ScaleData(sp_seu)
  sp_seu <- RunPCA(sp_seu)
}
if (!"umap" %in% names(sp_seu@reductions)) {
  sp_seu <- FindNeighbors(sp_seu, dims = 1:20)
  sp_seu <- FindClusters(sp_seu, resolution = 0.4)
  sp_seu <- RunUMAP(sp_seu, dims = 1:20)
}

# Show object context early
dim(sp_seu)
sp_seu
head(sp_seu@meta.data[, intersect(c("sample", "group", "region", "x", "y", "imagecol", "imagerow"), colnames(sp_seu@meta.data))])
table(Idents(sp_seu))
DimPlot(sp_seu, reduction = "umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE, pt.size = 0.45)
DimPlot(sp_seu, reduction = "umap", group.by = "region", pt.size = 0.45)
```

## 2) Score signatures

```{r spatial-signature-score, eval=requireNamespace("Seurat", quietly = TRUE)}
sp_genes <- rownames(sp_seu)
half_n <- max(30, floor(length(sp_genes) / 2))
idx_a <- seq_len(min(half_n, length(sp_genes)))
idx_b <- seq(from = min(half_n + 1, length(sp_genes)), to = length(sp_genes))
gs <- list(
  Spatial_signature_A = unique(sp_genes[idx_a])[1:min(30, length(unique(sp_genes[idx_a])))],
  Spatial_signature_B = unique(sp_genes[idx_b])[1:min(30, length(unique(sp_genes[idx_b])))]
)
sp <- score_signature(
  object = sp_seu,
  geneset = gs,
  geneset_source = "list",
  seurat = TRUE,
  assay = "RNA",
  layer = "data",
  slot = "data",
  method = "rank",
  min_genes = 3
)
```

## 3) Spatial score plots (slice + coordinates)

```{r spatial-slice-view, eval=requireNamespace("Seurat", quietly = TRUE), fig.width=7.5, fig.height=5.4, out.width="90%"}
coords <- if (all(c("x", "y") %in% colnames(sp_seu@meta.data))) {
  data.frame(x = sp_seu@meta.data$x, y = sp_seu@meta.data$y, row.names = rownames(sp_seu@meta.data))
} else if (all(c("imagecol", "imagerow") %in% colnames(sp_seu@meta.data))) {
  data.frame(x = sp_seu@meta.data$imagecol, y = sp_seu@meta.data$imagerow, row.names = rownames(sp_seu@meta.data))
} else {
  emb <- Embeddings(sp_seu, reduction = "pca")[, 1:2, drop = FALSE]
  data.frame(x = emb[, 1], y = emb[, 2], row.names = rownames(emb))
}

# Simulated tissue-like background for slice-context demonstration
tissue_bg <- as.raster(matrix(colorRampPalette(c("#f7f3e8", "#eadfca", "#d9c7a4"))(256), nrow = 16))

p_base <- plot_spatial_score(
  sp,
  pathway = rownames(sp$score)[1],
  coords = coords,
  image = tissue_bg,
  split.by = "sample"
)
p_base
```

## 4) Spatial differential testing

```{r spatial-differential, eval=requireNamespace("Seurat", quietly = TRUE), fig.width=7.5, fig.height=5.4, out.width="90%"}
sp_res <- test_signature(
  score = sp,
  region = "region",
  group = "group",
  sample = "sample",
  level = "sample_region",
  method = "wilcox"
)

top_sp <- sp_res$table$pathway[order(sp_res$table$p_adj)][1]
p_region <- plot_spatial_score(sp, pathway = top_sp, coords = coords, image = tissue_bg, split.by = "region")
p_cmp <- plot_spatial_compare(sp_res)

if (requireNamespace("patchwork", quietly = TRUE)) {
  p_region + p_cmp + patchwork::plot_layout(ncol = 2)
} else {
  p_region
  p_cmp
}
```
