---
title: "GLEAM full spatial transcriptomics workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GLEAM full spatial transcriptomics workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7.4,
  fig.height = 5.3,
  out.width = "88%",
  dpi = 120
)
library(GLEAM)
```

## 1) Spatial object construction

This workflow loads Seurat spatial objects, then combines expression and metadata
at matrix level to avoid image-slot compatibility breakpoints across Seurat
versions while retaining true spatial coordinate analysis.

```{r full-spatial-object, eval=requireNamespace("Seurat", quietly = TRUE), fig.width=7.6, fig.height=5.6, out.width="92%"}
library(Seurat)

a1_path <- system.file("extdata", "full_examples", "stxBrain_anterior1_seurat.rds", package = "GLEAM")
p1_path <- system.file("extdata", "full_examples", "stxBrain_posterior1_seurat.rds", package = "GLEAM")
if (a1_path == "") a1_path <- file.path("inst", "extdata", "full_examples", "stxBrain_anterior1_seurat.rds")
if (p1_path == "") p1_path <- file.path("inst", "extdata", "full_examples", "stxBrain_posterior1_seurat.rds")

sp_a1 <- readRDS(a1_path)
sp_p1 <- readRDS(p1_path)

resolve_expr <- function(obj) {
  assay_candidates <- unique(c(
    tryCatch(SeuratObject::DefaultAssay(obj), error = function(e) NULL),
    "Spatial",
    "RNA"
  ))
  assay_candidates <- assay_candidates[!is.na(assay_candidates) & nzchar(assay_candidates)]

  get_if_nonempty <- function(expr) {
    m <- tryCatch(eval.parent(substitute(expr)), error = function(e) NULL)
    if (!is.null(m) && nrow(m) > 0L && ncol(m) > 0L) return(m)
    NULL
  }

  for (assay in assay_candidates) {
    m <- get_if_nonempty(SeuratObject::LayerData(object = obj, assay = assay, layer = "data"))
    if (!is.null(m)) return(m)
    m <- get_if_nonempty(SeuratObject::LayerData(object = obj, assay = assay, layer = "counts"))
    if (!is.null(m)) return(m)
    m <- get_if_nonempty(SeuratObject::GetAssayData(object = obj, assay = assay, slot = "data"))
    if (!is.null(m)) return(m)
    m <- get_if_nonempty(SeuratObject::GetAssayData(object = obj, assay = assay, slot = "counts"))
    if (!is.null(m)) return(m)
  }

  stop("Failed to extract non-empty expression matrix from Seurat spatial object.")
}

prep_spatial_object <- function(obj, sample_label) {
  expr <- resolve_expr(obj)
  md <- as.data.frame(obj[[]], stringsAsFactors = FALSE)
  if (is.null(rownames(md))) rownames(md) <- colnames(expr)
  md <- md[colnames(expr), , drop = FALSE]
  if (!"sample" %in% colnames(md)) md$sample <- sample_label
  if (!"section" %in% colnames(md)) md$section <- sample_label
  if (!all(c("x", "y") %in% colnames(md))) {
    if (all(c("imagecol", "imagerow") %in% colnames(md))) {
      md$x <- md$imagecol
      md$y <- md$imagerow
    } else if (all(c("col", "row") %in% colnames(md))) {
      md$x <- md$col
      md$y <- md$row
    }
  }
  prefixed_ids <- make.unique(paste0(sample_label, "_", colnames(expr)), sep = "_dup")
  colnames(expr) <- prefixed_ids
  rownames(md) <- prefixed_ids
  list(expr = expr, meta = md)
}

sp1 <- prep_spatial_object(sp_a1, "anterior")
sp2 <- prep_spatial_object(sp_p1, "posterior")
common_genes <- intersect(rownames(sp1$expr), rownames(sp2$expr))
sp_expr <- cbind(sp1$expr[common_genes, , drop = FALSE], sp2$expr[common_genes, , drop = FALSE])
sp_md <- rbind(sp1$meta[colnames(sp1$expr), , drop = FALSE], sp2$meta[colnames(sp2$expr), , drop = FALSE])
sp_md <- sp_md[colnames(sp_expr), , drop = FALSE]
if (ncol(sp_expr) > 8000) {
  keep <- colnames(sp_expr)[1:8000]
  sp_expr <- sp_expr[, keep, drop = FALSE]
  sp_md <- sp_md[keep, , drop = FALSE]
}
if (!"group" %in% colnames(sp_md)) sp_md$group <- ifelse(grepl("anterior", sp_md$sample, ignore.case = TRUE), "anterior", "posterior")
if (!"region" %in% colnames(sp_md)) sp_md$region <- if ("seurat_clusters" %in% colnames(sp_md)) paste0("cluster_", sp_md$seurat_clusters) else sp_md$group
if (!"section" %in% colnames(sp_md)) sp_md$section <- sp_md$sample
if (!all(c("x", "y") %in% colnames(sp_md))) {
  if (all(c("array_col", "array_row") %in% colnames(sp_md))) {
    sp_md$x <- sp_md$array_col
    sp_md$y <- sp_md$array_row
  } else {
    n <- nrow(sp_md)
    sp_md$x <- seq_len(n)
    sample_vec <- if ("sample" %in% colnames(sp_md)) sp_md$sample else rep("sample_1", n)
    sp_md$y <- as.numeric(as.factor(sample_vec))
  }
}
if (!"pseudotime" %in% colnames(sp_md)) {
  sp_md$pseudotime <- rank(sp_md$x, ties.method = "average") / nrow(sp_md)
}
if (!"lineage" %in% colnames(sp_md)) {
  sp_md$lineage <- as.character(sp_md$region)
}

# Show object context before scoring
dim(sp_a1); sp_a1
dim(sp_p1); sp_p1
head(sp_md[, intersect(c("sample", "group", "region", "section", "x", "y", "imagecol", "imagerow"), colnames(sp_md))])
if ("umap" %in% names(sp_a1@reductions)) {
  DimPlot(sp_a1, reduction = "umap", group.by = "orig.ident", pt.size = 0.35)
}
```

## 2) Geneset scoring in spatial context

```{r full-spatial-score, eval=requireNamespace("Seurat", quietly = TRUE)}
sp_genes <- rownames(sp_expr)
half_n <- max(30, floor(length(sp_genes) / 2))
idx_a <- seq_len(min(half_n, length(sp_genes)))
idx_b <- seq(from = min(half_n + 1, length(sp_genes)), to = length(sp_genes))
gs <- list(
  Spatial_signature_A = unique(sp_genes[idx_a])[1:min(30, length(unique(sp_genes[idx_a])))],
  Spatial_signature_B = unique(sp_genes[idx_b])[1:min(30, length(unique(sp_genes[idx_b])))]
)

sp <- score_signature(
  expr = sp_expr,
  meta = sp_md,
  geneset = gs,
  geneset_source = "list",
  seurat = FALSE,
  method = "rank",
  min_genes = 3
)
```

## 3) Spatial visualization (true spatial context)

```{r full-spatial-map, eval=requireNamespace("Seurat", quietly = TRUE), fig.width=7.6, fig.height=5.6, out.width="92%"}
coords <- data.frame(x = sp_md$x, y = sp_md$y, row.names = rownames(sp_md))

# Slice/tissue-like background (replace with true tissue image when available)
tissue_bg <- as.raster(matrix(colorRampPalette(c("#f8f5ea", "#e8dcc4", "#d2b48c"))(256), nrow = 16))

plot_spatial_score(sp, pathway = rownames(sp$score)[1], coords = coords, image = tissue_bg)
plot_spatial_score(sp, pathway = rownames(sp$score)[2], coords = coords, split.by = "section", image = tissue_bg)
plot_spatial_multi(sp, pathways = rownames(sp$score)[1:4], coords = coords)
```

## 4) Spatial differential analysis

```{r full-spatial-diff, eval=requireNamespace("Seurat", quietly = TRUE), fig.width=7.6, fig.height=5.6, out.width="92%"}
# region vs region
res_region <- test_signature(sp, region = "region", level = "region", method = "wilcox")

# group vs group within region
res_group_region <- test_signature(
  sp,
  region = "region",
  group = "group",
  sample = "sample",
  level = "sample_region",
  method = "wilcox"
)

top_sp <- res_group_region$table$pathway[order(res_group_region$table$p_adj)][1]
plot_spatial_score(sp, pathway = top_sp, coords = coords, image = tissue_bg, split.by = "region")
plot_spatial_compare(res_group_region)
```

## 5) Optional embedding linkage and trajectory-like display

```{r full-spatial-embedding-link, eval=requireNamespace("Seurat", quietly = TRUE), fig.width=7.6, fig.height=5.6, out.width="92%"}
emb2 <- as.matrix(sp_md[, c("x", "y"), drop = FALSE])
rownames(emb2) <- rownames(sp_md)
colnames(emb2) <- c("Spatial_1", "Spatial_2")
plot_embedding_score(sp, pathway = rownames(sp$score)[1], embedding = emb2, reduction = "umap")
plot_pseudotime_score(sp, pathway = rownames(sp$score)[1], pseudotime = "pseudotime", lineage = "lineage")
```
