---
title: "GLEAM quick start: scRNA-seq Seurat workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GLEAM quick start: scRNA-seq Seurat workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7.2,
  fig.height = 5.2,
  out.width = "85%",
  dpi = 120
)
library(GLEAM)
```

## Goal

This quick start demonstrates a practical Seurat-first scRNA workflow:

1. Start from a Seurat object
2. Run standard preprocessing (normalization, HVG, PCA, neighbors, clustering, UMAP/tSNE)
3. Load IFNB Seurat example object
4. Load Hallmark gene sets
5. Score signature activity
6. Run differential analysis after scoring
7. Visualize scores with patchwork composition

## 1) Prepare Seurat object and run standard preprocessing

Before signature scoring, we run a conventional Seurat preprocessing path so that:

- embeddings (`umap`, `pca`, `tsne`) are available for score visualization;
- cluster labels are available for state-level interpretation;
- downstream differential comparisons are based on a well-structured object.

```{r seurat-object, eval=requireNamespace("Seurat", quietly = TRUE), fig.width=7.5, fig.height=5.4, out.width="90%"}
library(Seurat)

# Use full IFNB Seurat example object
ifnb_path <- system.file("extdata", "full_examples", "ifnb_seurat.rds", package = "GLEAM")
if (ifnb_path == "") ifnb_path <- file.path("inst", "extdata", "full_examples", "ifnb_seurat.rds")
seu <- readRDS(ifnb_path)
if (ncol(seu) > 5000) seu <- subset(seu, cells = colnames(seu)[1:5000])

md <- seu@meta.data
if (!"sample" %in% colnames(md)) md$sample <- if ("orig.ident" %in% colnames(md)) as.character(md$orig.ident) else "sample_1"
if (!"group" %in% colnames(md)) md$group <- if ("stim" %in% colnames(md)) as.character(md$stim) else ifelse(seq_len(nrow(md)) <= nrow(md)/2, "A", "B")
if (!"celltype" %in% colnames(md)) md$celltype <- if ("seurat_clusters" %in% colnames(md)) paste0("cluster_", md$seurat_clusters) else as.character(Idents(seu))
seu@meta.data <- md

# Minimal QC context for a quick start (no aggressive filtering here).
if (any(grepl("^MT-", rownames(seu)))) {
  seu[["percent.mt"]] <- PercentageFeatureSet(seu, pattern = "^MT-")
}
qc_cols <- intersect(c("nFeature_RNA", "nCount_RNA", "percent.mt"), colnames(seu@meta.data))
summary(seu@meta.data[, qc_cols, drop = FALSE])

if (!"pca" %in% names(seu@reductions)) {
  seu <- NormalizeData(seu)
  seu <- FindVariableFeatures(seu)
  seu <- ScaleData(seu)
  seu <- RunPCA(seu)
}
if (!"umap" %in% names(seu@reductions)) {
  seu <- FindNeighbors(seu, dims = 1:20)
  seu <- FindClusters(seu, resolution = 0.4)
  seu <- RunUMAP(seu, dims = 1:20)
}
if (!"tsne" %in% names(seu@reductions)) {
  seu <- RunTSNE(seu, dims = 1:20)
}

# Show object context early
dim(seu)
seu
head(seu@meta.data[, c("sample", "group", "celltype")])
table(Idents(seu))
DimPlot(seu, reduction = "umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE, pt.size = 0.4)
DimPlot(seu, reduction = "umap", group.by = "celltype", pt.size = 0.4)
```

## 2) Load built-in genesets

```{r eval=requireNamespace("Seurat", quietly = TRUE)}
map_geneset_to_expr <- function(gs, expr_genes) {
  expr_genes <- as.character(expr_genes)
  expr_upper <- toupper(expr_genes)
  mapped <- lapply(gs, function(g) {
    idx <- match(toupper(unique(as.character(g))), expr_upper, nomatch = 0L)
    unique(expr_genes[idx[idx > 0L]])
  })
  mapped[vapply(mapped, length, integer(1)) > 0L]
}

fallback_signatures <- function(expr_genes) {
  g <- unique(as.character(expr_genes))
  g <- g[!is.na(g) & nzchar(g)]
  k <- min(30L, max(3L, floor(length(g) / 4)))
  idx1 <- seq_len(min(k, length(g)))
  idx2 <- seq.int(max(1L, length(g) - k + 1L), length(g))
  list(
    Signature_A = g[idx1],
    Signature_B = g[idx2]
  )
}

gs_hallmark <- NULL
for (sp in c("human", "mouse")) {
  gs_try <- try(get_geneset("hallmark", source = "builtin", species = sp), silent = TRUE)
  if (inherits(gs_try, "try-error")) next
  gs_try <- map_geneset_to_expr(gs_try, rownames(seu))
  gs_try <- gs_try[vapply(gs_try, length, integer(1)) >= 3L]
  if (length(gs_try) > 0L) {
    gs_hallmark <- gs_try
    break
  }
}
if (is.null(gs_hallmark) || length(gs_hallmark) == 0L) {
  message("[GLEAM] No matched Hallmark signatures; using object-derived fallback signatures.")
  gs_hallmark <- fallback_signatures(rownames(seu))
}
```

## 3) Score signatures

```{r score-signature, eval=requireNamespace("Seurat", quietly = TRUE)}
sc <- score_signature(
  object = seu,
  geneset = gs_hallmark,
  geneset_source = "list",
  seurat = TRUE,
  assay = "RNA",
  layer = "data",
  slot = "data",
  method = "ensemble",
  min_genes = 3
)
```

## 4) Differential analysis after scoring

```{r signature-differential, eval=requireNamespace("Seurat", quietly = TRUE)}
res <- test_signature(
  score = sc,
  group = "group",
  sample = "sample",
  celltype = "celltype",
  level = "pseudobulk",
  method = "wilcox"
)
```

## 5) Core visualizations (patchwork layout)

```{r quickstart-scrna-plots, eval=requireNamespace("Seurat", quietly = TRUE), fig.width=7.5, fig.height=5.4, out.width="90%"}
top_pw <- res$table$pathway[order(res$table$p_adj)][1]
p1 <- plot_embedding_score(sc, pathway = top_pw, object = seu, reduction = "umap")
p2 <- plot_embedding_score(sc, pathway = top_pw, object = seu, reduction = "pca")
p3 <- plot_violin(sc, pathway = rownames(sc$score)[1], group = "group")

if (requireNamespace("patchwork", quietly = TRUE)) {
  p1 + p2 + p3 + patchwork::plot_layout(ncol = 3)
} else {
  p1
  p2
  p3
}
```
