---
title: "GLEAM quick start: scRNA-seq Seurat workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GLEAM quick start: scRNA-seq Seurat workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(GLEAM)
```

## Goal

This quick start demonstrates a practical Seurat-first scRNA workflow:

1. Start from a Seurat object
2. Load Hallmark and KEGG gene sets
3. Score pathway activity
4. Run differential analysis after scoring
5. Visualize scores and differential outputs with Seurat-style displays

## 1) Prepare Seurat object

```{r, eval=FALSE}
library(Seurat)

# Build a lightweight demo object from bundled matrix+metadata
data("pbmc_medium_matrix", package = "GLEAM")
data("pbmc_medium_meta", package = "GLEAM")
seu <- CreateSeuratObject(counts = pbmc_medium_matrix, meta.data = pbmc_medium_meta)

seu <- NormalizeData(seu)
seu <- FindVariableFeatures(seu)
seu <- ScaleData(seu)
seu <- RunPCA(seu)
seu <- RunUMAP(seu, dims = 1:20)
seu <- RunTSNE(seu, dims = 1:20)
```

## 2) Load built-in and curated genesets

```{r, eval=FALSE}
gs_hallmark <- get_geneset("hallmark", source = "builtin", species = "human")
gs_kegg <- get_geneset(NULL, source = "kegg", species = "human") # msigdbr optional
```

## 3) Score pathways

```{r, eval=FALSE}
sc <- score_pathway(
  object = seu,
  geneset = gs_hallmark,
  geneset_source = "list",
  seurat = TRUE,
  assay = "RNA",
  layer = "data",
  slot = "data",
  method = "ensemble"
)
```

## 4) Differential analysis after scoring

```{r, eval=FALSE}
res <- test_pathway(
  score = sc,
  group = "group",
  sample = "sample",
  celltype = "celltype",
  level = "pseudobulk",
  method = "wilcox"
)
```

## 5) Core visualizations

```{r, eval=FALSE}
# FeaturePlot-like embedding displays
top_pw <- res$table$pathway[order(res$table$p_adj)][1]
plot_embedding_score(sc, pathway = top_pw, object = seu, reduction = "umap")
plot_embedding_score(sc, pathway = top_pw, object = seu, reduction = "pca")
plot_embedding_score(sc, pathway = top_pw, object = seu, reduction = "tsne")

# Dot-bar and distribution displays
plot_dot_bar(sc, by = c("group", "celltype"))
plot_violin(sc, pathway = rownames(sc$score)[1], group = "group")
plot_ridge(sc, pathway = rownames(sc$score)[1], group = "celltype")

# Differential summary
plot_volcano(res)
```
